A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-19 paper/bare_conf.tex
Add a bit of code transformations

\author{\IEEEauthorblockN{Fabio Luporini}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: f.luporini12@imperial.ac.uk}
\and
\IEEEauthorblockN{Paul H. J. Kelly}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: p.kelly@imperial.ac.uk}
\and
\IEEEauthorblockN{David Ham}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: d.ham@imperial.ac.uk}}

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Burgers problem (3D mesh, Lagrange $p=1$ elements).}
\label{code:burgers}
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//It follows 11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//It follows 10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[ip][r]);\\
~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K[5]*F9)+(K[8]*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K[0]*X1$\_$D100[i][k])+(K[3]*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K[6]*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K[2]*X2$\_$D100[i][k])+...+(K[8]*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K[2]*X2$\_$D100[i][j])+...+(K[8]*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$a hundred of sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\end{algorithm}

@@

\section{Code Transformations}
\label{sec:code-transf}
%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead. 
%Follows a description of code transformations using our compiler-driven approach.

@@

%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{Pictures/helmholtz-normalized.pdf}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{Pictures/advdiff-normalized.pdf}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}

>>>>

\author{\IEEEauthorblockN{Fabio Luporini}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: f.luporini12@imperial.ac.uk}
\and
\IEEEauthorblockN{Florian Rathgeber}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: f.rathgeber10@imperial.ac.uk}
\and
\IEEEauthorblockN{David Ham}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: d.ham@imperial.ac.uk}}

%\and
%\IEEEauthorblockN{Paul H. J. Kelly}
%\IEEEauthorblockA{Department of Computing\\
%Imperial College London\\
%London, UK\\
%Email: p.kelly@imperial.ac.uk}

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Burgers problem (3D mesh, Lagrange $p=1$ elements).}
\label{code:burgers}
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//It follows 11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//It follows 10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[ip][r]);\\
~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K[5]*F9)+(K[8]*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K[0]*X1$\_$D100[i][k])+(K[3]*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K[6]*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K[2]*X2$\_$D100[i][k])+...+(K[8]*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K[2]*X2$\_$D100[i][j])+...+(K[8]*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred of sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\end{algorithm}

@@

\section{Domain-Driven Code Transformations}
\label{sec:code-transf}
%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead. 
%Follows a description of code transformations using our compiler-driven approach.

% Autovectorizable LICM -> show code snippet?
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion ($licm$). Vendor compilers apply $licm$, although not in the systematic way we need in our assembly kernels. We want to overcome two deficiencies that both $icc$ and $gcc$ have. First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$ only, which are not automatically lifted. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations by doing source-level $licm$. In addition, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension. This is implemented by introducing temporary arrays and new loops in the nest. At the price of both extra memory for storing temporaries and reduced register locality, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which can be otherwise significant.

% Padding and Data Alignment
Auto-vectorization of assembly code that computes $A$ can be less effective if data are not aligned and if the length of the innermost loop is small compared to the hardware vector length ($vl$). Data alignment is enforced in two steps. Initially, both arrays and matrices are allocated to addresses that are multiples of $vl$. Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$. For example, assume the original size of a matrix is 3$\times$3, and the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use double-precision floating-point values (64 bits). Then, the padded matrix on this architecture is 3$\times$4. Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$. This avoids the introduction of a remainder (scalar) loop from the compiler, which would be responsible for inefficient vectorization.

% Register tiling in its two forms
One notable problem assembly kernels are exposed to concerns register allocation and locality. There exist two critical situations that can negatively impact performance. The first occurs when trip counts are sufficiently large to prevent register reuse along the innermost dimension ($k$). A combination of the two scenarios is also possible. We want to exploit the small size of the iteration space and the structure of the computation to optimize register management.

% remember: op vectorization PLUS unroll and jam !!

% Splitting

% mention (obvious) composability

@@

% We fix a specific loop permutation (i esterno perche minimizza.... swapping j or k, after LICM, is equivalent)
